---
title: "03: Importing data"
subtitle: "Introduction to R for health data"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
editor: visual
---

------------------------------------------------------------------------

## Content

-   Importing and exporting data files
    -   csv
    -   other delimited files
    -   rds
    -   xlsx
    -   ods
    -   dta
-   Reading and writing on a SQL database

------------------------------------------------------------------------

## Recap from 02_visualise

1.  Explore the in-built dataset below.
    -   This contains data from an experiment on theophylline, a drug used to treat asthma and COPD.
    -   Each subject was given a dose of the drug and then had repeat blood samples taken to measure its serum concentration.

```{r}
theoph <- tibble::as_tibble(datasets::Theoph)

# Your code goes here ↓

```

2.  Create a concentration-time graph with `ggplot2`:
    -   time from dose will be on the x-axis
    -   concentration will be on the y-axis
    -   there will be a coloured line indicating each participant, showing how their concentration changed through time
    -   *bonus*: also add a point for each (time, concentration) value.

```{r}

```

------------------------------------------------------------------------

## Load packages

In this session, you'll use functions from a few packages:

-   [`readr`](https://readr.tidyverse.org/reference/index.html) to read and write files in most formats - part of the core tidyverse, so it's automatically loaded when calling `library(tidyverse)`
-   [`readxl`](https://readxl.tidyverse.org/) to read Excel spreadsheets
-   [`haven`](https://haven.tidyverse.org/) to read and write dta files
-   [`readODS`](https://docs.ropensci.org/readODS/) to read and write ODS spreadsheets
-   `DBI`, `odbc` and to read data from (and write it to) an SQL database
-   [`dbplyr`](https://dbplyr.tidyverse.org/) to simplify querying a SQL database table.
-   [`fs`](https://fs.r-lib.org/) to handle file and folders

Before running this chunk, please ensure that you have these packages installed.

```{r}
library(tidyverse)
library(haven)
library(readxl)
library(readODS)
library(DBI)
library(odbc)
library(dbplyr)
library(fs)
```

------------------------------------------------------------------------

## Importing and exporting data files

### Importing data from a file

The general process is this:

1.  You choose a reading function that matches the file type; for example, `readr::read_csv()` for CSV files.
2.  You list the path to your data as an argument of that function; for example, `read_csv("data/lab.csv")`.
3.  You assign the output of this function to an object with a name of your choice; for example, `lab_raw <- read_csv("data/lab.csv")`.

#### Importing csv files

CSV files (comma-separated values) are the most common file type for tabular data.

I recommend using function `read_csv()` (note the underscore) from package `readr`, rather than the base R function `read.csv`. This is because it's faster, among other things.

You need to list the path to the file as the first argument of this function.

```{r}
metabolic <- read_csv("data/raw/metabolic.csv")
```

The code above will work if the CSV `metabolic.csv` is stored in a folder called `data` which is kept where this .qmd script is.

However, if you need to keep your data on a shared drive, you need to specify the longer path. I recommend using the full path to identify the data folder (e.g., `//FILECOL17.phe.gov.uk/ProjectData/HCAI/intRo/data/`) and not just e.g. `H:/intRo/data/`, because some people might not have mapped the same drive to the same letter.

You can then use `paste0()` or `glue::glue()` to identify your file. Please note, I recommend **against** using `setwd()` to do this. More information on data location will be covered in session `13_workflows`.

------------------------------------------------------------------------

You will note that now this dataset has been imported in your *Environment*, in the top right of your window.

Let's preview it.

```{r}
head(metabolic)
```

##### Exercise 1

1.  Choose any other CSV file in the /data folder
2.  Import it.
3.  Preview it with `head()` or another method used in session `02_visualise`.

```{r}

```

Please note that if the column headers contain spaces, `read_csv` will keep them. In session `08_cleaning` we'll go through how to make these column names easier to deal with.

------------------------------------------------------------------------

If the file contains many columns that you don't need, you can specify which ones to import with the `col_select` argument. This has the advantage of speeding up the import.

```{r}
smoking <- read_csv("data/raw/metabolic.csv", col_select = c(patient_id, smoker))
```

`read_csv()` will also guess which data type each column contains: dates, characters, integers, double (i.e., real numbers), etc. This guess is based on the first 1000 rows. You can change this default by setting argument `guess_max` to a different number, or you can specify the type of each column:

```{r}
ages <- read_csv(
    "data/raw/metabolic.csv",
    col_select = c(patient_id, age),
    col_types = c(col_character(), col_integer())
    )
```

`read_csv` also has many other arguments; for example, you can use `skip` to avoid reading some initial rows that don't contain data, or `n_max` to establish the maximum number of rows to read (useful if you just want to preview a data file, for example). Run `help(read_csv)` for more information.

#### Importing other delimited files

You can use `readr::read_delim()`, which is very similar to `read_csv`. You'll need to specify which delimiter the files uses: for example `read_delim(..., delim = "|")`.

#### Even faster importing

If you need to import a very large delimited file, you could try:

-   only selecting the columns that you will need (see `col_select` above)
-   using `data.table::fread()` instead of the `readr` functions.

NB: if you're reading a file from a shared drive that you're accessing remotely, the speed bottleneck will likely be your internet download speed, not the reading function you pick.

Very large files should ideally be saved in a different format that makes it easier to import, for example .fst.

#### Importing rds files

R's native data file format, rds, keeps data types, so it's helpful to use for example for mid-processing files.

You can read them with `readr::read_rds()` or base R `readRDS()` :

##### Exercise 2

Read `diabetes.rds`, a file saved in the `/data/processing/` folder.

```{r}

```

#### Importing Excel files

You can import Excel spreadsheets with `readxl::read_excel()`, which will work with both the old .xls and the new .xlsx formats.

```{r}
candi_site5 <- read_excel("data/raw/candidaemias.xlsx")
```

Please note that by default `read_excel()` imports the first sheet, but you can specify otherwise.

##### Exercise 3

1.  Open the documentation of `read_excel()` .
2.  Find out how to import a different sheet.
3.  Read the sheet containing antifungal treatment from the `candidaemias.xlsx` file.

```{r}

```

#### Importing ODS files

You can use `readODS::read_ods()`, which is very similar to the function above.

#### Importing dta files

If for any unpleasant circumstance the data was saved in Stata's proprietary dta format, `haven::read_dta()` comes to the rescue.

##### Exercise 4

1.  Import the `point_prevalence.dta` file.
2.  Preview the data. What do you notice about the `site`, `site_type` and `fy` columns?

```{r}

```

Any columns containing Stata-labelled values will be imported as a special class called `labelled`. You should convert these to factor shortly after importing, by using `haven::as_factor()` as shown in the code below. For the moment, please don't focus on the `|>`, `mutate`, `across` and `where` bits: we'll cover these in future sessions.

```{r}
pps <-
    read_dta("data/processing/point_prevalence.dta") |> 
    # Make labelled values readible
    mutate(across(where(is.labelled), as_factor))

pps
```

------------------------------------------------------------------------

### Exporting data to a file

Whatever you do to the dataset, it doesn't change the original file. If you want to make some changes permanent, you need to use a writing function.

Important: never overwrite the initial, raw data file. Ideally, separate the data in different folders according to their type: raw files, processing tables, processed tables. Let's create a new folder using `fs::dir_create()`.

```{r}
dir_create("data/processed")
```

#### Exporting as csv

You can use `readr::write_csv`. For example, let's save the pps data frame in CSV format.

```{r}
write_csv(pps, "data/processed/point_prevalence.csv")
```

#### Exporting as rds

```{r}

```

`readr::write_rds` or base R `saveRDS()` let you save in the native R format, which will keep the data types. I recommend doing this for intermediate data tables you are processing.

##### Exercise 5

Save the `pps` data frame as an RDS file in the "processing" folder.

```{r}

```

I recommend *against* saving to .RData because it doesn't make it clear which objects are imported in this way.

#### Exporting as xlsx

I don't recommend saving in xlsx or xls format, but if you need to, you can use [`writexl`](https://docs.ropensci.org/writexl/)`::write_xlsx()` .

#### Exporting as ODS

If you need to export data in spreadsheet format, I recommend using ODS (OpenDocument spreadsheet), as it has an open licence. You can do this with `readODS::write_ods()`.

#### Exporting as dta

If for any unfortunate circumstance you must save the data in Stata's proprietary dta format, you can use `haven::write_dta()`.

------------------------------------------------------------------------

## Reading and writing on a SQL database

### Initial SQL setup

The very first time that you use a SQL database, you'll need to:

1.  Get access to that database
2.  Add the data source to your ODBC Data Sources

The code in the sections below assumes that you have a Data Lake source in your ODBC Data Sources.

### Connect to a SQL database

Now we need to connect to the database. You can do this directly from R, using functions from package `DBI` and `odbc`.

`dbConnect` has many options and choosing which to use depends on your database - you should ask other data scientists in your team how to set up a connection.

If you only need to import tables from a single database, you can specify the database in this connection.

```{r}
con_LookupsShared <-
  dbConnect(
    odbc(),
    Driver = "SQL Server",
    Server = "SQLClusColLK19\\Lake19",
    Database = "LookupsShared"
    )
```

If the code in the chunk above worked, you will now see that the Connections tab in the panel at the top right of this window will have become active and show all databases within this server.

##### Exercise 6

Create a separate connection to another database within a server you have access to.

```{r}

```

### Pull data from a SQL table using `dbplyr`

You can get data from a SQL database table in a variety of ways. In this topic, we'll use package `dbplyr` to query a database using `dplyr` code - detailed below. (In topic 10_sql, we'll explore some other methods.)

1.  Use `tbl()` to create an object that represents the table you want. NB - running this code will not import the whole table into memory.

```{r}
sicbl_db <- tbl(con_LookupsShared, "vLKP_SICBL24")
```

2.  Use `dplyr` functions to query the database. This will be covered in session `06_subset`, but for now just be aware that `select` keeps only the columns listed. Be aware that this still is not a dataframe, it's just a SQL command. In the background, `dbplyr` is converting the `dplyr` code into a SQL query. You can see the SQL query by using `show_query()` on the new object.

```{r}
sicbl_name_lookup_db <- select(sicbl_db, SICBL24CD, SICBL24NM)

show_query(sicbl_name_lookup_db)
```

3.  Use `collect` to run the query and return the data into a tibble.

```{r}
sicbl_name_lookup <- collect(sicbl_name_lookup_db)
sicbl_name_lookup
```

------------------------------------------------------------------------

### Writing data to a SQL database

If you need to write data to SQL, you can use `DBI::dbWriteTable()`.

Imagine you want to upload a data.frame called `test` to the LookupsShared database. You need to list the connection object, the name you want the table to have on the database, and the data.frame object name.

Please note that this chunk will throw an error - you don't have writing permission for this database.

```{r}
test <- tibble(col1 = c("a", "b", "c"))

dbWriteTable(
  con_LookupsShared,
  name = "vLKP_test",
  value = test
  )
```

------------------------------------------------------------------------

##### Stretch exercise

If you have access to LookupsShared:

1.  Find the table containing the 2013 European Standard Population
2.  Create an object that represents this table
3.  Read all the data in this table and store it in a `data.frame`
4.  Visualise this data with a bar chart.

```{r}

```

------------------------------------------------------------------------

Once you're done with a SQL database, please disconnect from it.

```{r}
dbDisconnect(con_LookupsShared)
```

------------------------------------------------------------------------

## Learning more

I recommend these resources:

-   Data import with the tidyverse [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-import.pdf)
-   R for Data Science (2e)
    -   (csv) [Data import](https://r4ds.hadley.nz/data-import.html)
    -   [Spreadsheets](https://r4ds.hadley.nz/spreadsheets)
    -   [Databases](https://r4ds.hadley.nz/databases)

------------------------------------------------------------------------

## Solutions

```{r}
#| label: recap Q1
theoph
```

```{r}
#| label: recap Q2
theoph |>
  ggplot(aes(x = Time, y = conc, colour = Subject)) +
  geom_line() +
  geom_point()
```

```{r}
#| label: exerc1 solution

ambi_raw <- read_csv("data/raw/ambition.csv")
glimpse(ambi_raw, width = 90)
```

```{r}
#| label: exerc2 solution

dm <- read_rds("data/processing/diabetes.rds")
```

```{r}
#| label: exerc3 solution

help(read_excel)
candi_antifung <- read_excel("data/raw/candidaemias.xlsx", sheet = "meds")
```

```{r}
#| label: exerc4 solution

pps_raw <- read_dta("data/processing/point_prevalence.dta")
pps_raw
```

```{r}
#| label: exerc5 solution

write_rds(pps, "data/processing/point_prevalence.rds")
```

```{r}
#| label: exerc6 solution

# Just an example - this will only work if you have access to this database
con_staging <-
  dbConnect(
    odbc(),
    Driver = "SQL Server",
    Server = "SQLCLUSCOLHFN17\\HFN17",
    Database = "HCAI_Analysis_Staging")
```

```{r}
#| label: stretch exerc solution

esp_db <- tbl(con_LookupsShared, "vRef_ESP2013")

esp <- collect(esp_db)

ggplot(esp, aes(Age_Band_Min, Population)) +
    # The data is already aggregated
    geom_col()
```
