---
title: "12: Quarto"
subtitle: "Introduction to R for health data"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
format:
  docx: 
    fig-dpi: 200
    reference-doc: templates/document_template.docx
editor: visual
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggtext)
source("scripts/functions.R")

bsi_resistance <- read_rds("data/processed/bsi_resistance.rds")
bsi_resistance
```

{{< pagebreak >}}

## Antibiotic resistance in bloodstream infection isolates

### *Klebsiella pneumoniae*

The percentage of *Klebsiella pneumoniae* isolates resistant to piperacillin with tazobactam increased from `r get_perc_r("K. pneumoniae", "piperacillin/tazobactam", 2018)` to `r get_perc_r("K. pneumoniae", "piperacillin/tazobactam", 2022)` between 2018 and 2022.

### *Acinetobacter* species

For *Acinetobacter* spp., the percentage of isolates resistant to gentamicin decreased from `r get_perc_r("Acinetobacter spp.", "gentamicin", 2018)` to `r get_perc_r("Acinetobacter spp.", "gentamicin", 2022)` between 2018 and 2022.

### All bacteria

```{r}
#| label: bar_chart
#| echo: false
#| fig-width: 6.3
#| fig-height: 6

bsi_resistance |> 
  ggplot(aes(x = year, y = proportion, fill = result)) +
  facet_grid(vars(organism_label), vars(antibiotic)) +
  geom_col() +
  scale_fill_manual(values = c("S/I" = "#00A5DF", "R" = "#E40046")) +
  scale_x_continuous(breaks = seq(2018, 2022, 2)) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_line(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    strip.text = element_markdown()
    ) +
  labs(
    fill = "Result",
    caption = "Data source: ESPAUR 2022 to 2023 report, tables 2.4 and 2.5"
  )
```

The underlying data tables are available on the [ESPAUR page on gov.uk](https://www.gov.uk/government/publications/english-surveillance-programme-antimicrobial-utilisation-and-resistance-espaur-report).

------------------------------------------------------------------------

## Learning more

-   R for Data Science (2ed)
    -   [Quarto](https://r4ds.hadley.nz/quarto.html)
    -   [Quarto formats](https://r4ds.hadley.nz/quarto-formats)
-   Quarto.org
    -   [Markdown basics](https://quarto.org/docs/authoring/markdown-basics.html)
    -   [R computations](https://quarto.org/docs/computations/r.html)
    -   [Inline code](https://quarto.org/docs/computations/inline-code.html)
    -   [Word basics](https://quarto.org/docs/output-formats/ms-word.html)
    -   [Word templates](https://quarto.org/docs/output-formats/ms-word-templates.html)
