---
title: "1: The basics"
subtitle: "Introduction to R"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
---

-------------------------------------------------------------------------------

# What you will learn

* User interface
* Console
* R Markdown notebooks
* Assignment
* Functions
* Datasets

-------------------------------------------------------------------------------

# User interface

The RStudio interface is divided into four parts:
`+--------------+-------------+`
`| Code,        | Environment |`
`| Viewer       |             |`
`|              |             |`
`+--------------+-------------+`
`| Console      | Files,      |`
`|              | Packages,   |`
`|              | Help        |`
`+--------------+-------------+`

- *top left*: the code script  or notebook you're working on (like this one: `intRo1 basics.Rmd`).
- *bottom left*: the console (where you can input your commands)
- *top right*: Environment (where your datasets and other "objects" are stored)
- *bottom right*: help (and other stuff)

-------------------------------------------------------------------------------

# Console

The console is used to run simple commands that you don't need to save.

*Exercise*
In the console, add 3 and 5. You do this by using it as a calculator: 3+5 Enter.

Now forget about the console – it's better to keep your code in a notebook (or in a script). This is for many reasons:

1. Easily editing your code when it doesn't work the first time you run it.
2. Keeping track of your work so you don't have to restart from scratch every time.
3. Sharing your work with others

-------------------------------------------------------------------------------

# R Notebooks (.Rmd)

A notebook, like this one (`intRo1 basics.Rmd`) is composed of two things:

* Text, which is the main component
* "Chunks", which contain code (and comments, which are preceded by #)

A chunk looks like this:
```{r}

```
Please note that those three signs are not normal apostrophes ('): they are "backticks" (`). On a UK/US Windows keyboard and US Mac keyboard, the key is to the left of the number 1. On a UK Mac keyboard, the key is to the left of Z.

To run code in a chunk, you click on the *Run* (green triangle) button at the top right of the chunk (or you place your cursor inside it and press *Ctrl+Shift+Enter* (on Mac: *Cmd+Option+Enter*).

*Exercise*
Run this chunk. Note where is the output displayed.
```{r}
5 - 1
```

To run just part of the chunk, you can select the code of interest (or just place your cursor on any line) and press *Ctrl+Enter* (on Mac: *Cmd+Enter*).

*Exercise*
Only run the second line of this chunk.
```{r}
# Area of two circles
pi * 4 ^2
pi * 5.7 ^2
```

To add a new chunk, you click on the toolbar: `Code` -> `Insert Chunk`.

*Exercise*
Add a new chunk in the space below, and use it to divide 6 by 2.




-------------------------------------------------------------------------------

# R scripts (.R)

An R script, for example `analysis.R`, mainly contains code, and any non-code text needs to be highlighted with a preceding `#`. The output will only appear in the console.

We won't be using R scripts in this course, but you might see them elsewhere.

To create a new R script, you go in the toolbar: `File` -> `New File` -> `R Script`, or with a keyboard short-cut: `Ctrl` + `Shift` + `N` on Windows.

-------------------------------------------------------------------------------

# Assignment

You can tell R that something has a name; R will remember it and you can use that later in your code. This is called *assignment* and is done by using the arrow <- (keyboard shortcut: `Alt` + `*` on Windows, `Option` + `*` on Mac) Assignment creates a new "object".
The following code means: "radius" gets the value of 5.7. "radius" is a new object.

*Exercise*
Run this chunk. In which part of the screen is R storing this value?
```{r}
# Assigning a value to "radius"
radius <- 5.7

# Visuaising the value of "radius"
radius
```

From now, you can use the object "radius" in the rest of your code, and R will know what you mean.

*Exercise*
Calculate the circumference of a circle of radius 5.7, without typing 5.7 (Formula: twice the radius times pi)
```{r}
# ...
```

Note that assigning a new value to an existing object will permanently replace the old version.
```{r}
radius <- 6.58
radius
```


-------------------------------------------------------------------------------

# Functions

In programming, a function is like a verb. It takes something (its "arguments"), it does something to it, and gives you back an output.
In R, the arguments of a function are written in brackets.
```{r}
# Square root function
sqrt(64)
```

*Exercise*
Create a new chunk below, and use it to calculate the square root of 6.58, without typing 6.58.



-------------------------------------------------------------------------------

# Datasets

R is mostly built around dataframes: data in table format.

Some dataframes are included in R so we can practice.

For example, let's explore `ToothGrowth`, a small dataset from an experiment on the impact of vitamin C supplementation on tooth length among 60 guinea pigs.

Calling just the name of the dataset will show us a preview:

To have a look at this dataset, you can call its name or you can use the function `View()`. Caution: R is case-sensitive!

```{r}
ToothGrowth
```

```{r}
# This will open a new tab!
View(ToothGrowth)
```

A dataframe has columns called *variables*. These are the variables in `ToothGrowth`:
* `len`: tooth cell length
* `supp`: supplement type (VC = vitamin C; OJ = orange juice)
* `dose`: dose of vitamin C in mg/day


## Using functions on a dataset column

Now let's use the `mean()` and `sd()` functions to calculate the average tooth cell length and its standard deviation.

R can work with more than one dataset at the time. For this reason, every time you refer to a variable, you also need to tell R which dataset is that variable from. You do this by typing a `$` sign between the dataset and the variable name.
(The `$` sign is technically called the "exposition operator").
```{r}
mean(ToothGrowth$len)
# The above reads: calculate the mean of a variable taken from `ToothGrowth` and called `len`.

sd(ToothGrowth$len)
```


*Exercise*
Let's use another dataset, `chickwts`. This contains info on chicken weights (in grams) and their feed type.

1. Calculate the `median` chicken weight.
2. Calculate the standard deviation of chicken weight.
```{r}
# Preview the dataframe
chickwts

# Median chicken weight
# ...

# Standard deviation
# ...

```

*Recap exercise*
1. Explore the `Theoph` dataset
2. Calculate the mean weight of its study participants
3. Convert this mean weight from kilograms to pounds. (1 kg = approx. 2.2 pounds)

-------------------------------------------------------------------------------

# Solutions

```{r}
# Second-to-last exercise
median(chickwts$weight)
sd(chickwts$weight)

# Recap exercise
# 1. Explore dataset
View(Theoph)

# 2. Calculate mean weight
mean(Theoph$Wt)

# 3. Convert mean weight to pounds
mean(Theoph$Wt) * 22

# An alternative way
mean_kg <- mean(Theoph$Wt)
mean_kg * 2.2
```

-------------------------------------------------------------------------------
