---
title: "1: The basics"
subtitle: "Introduction to R"
author: Andrea Mazzella[link](https://github.com/andreamazzella)
output: html_notebook
---

# What you will learn
a) Console
b) R Notebooks
c) Assignment
d) Functions
e) Datasets

# Console
The console is used to run simple commands that you don't need to save.

*Exercise*
In the console below, add 3 and 5 - it's like using a calculator: 3+5 Enter.

Now forget about the console forever.

# R Notebooks

A notebook is composed of two things:
a) Text, which is the main component
b) "Chunks", which contain code (plus comments, preceded by #)

A chunk is contained by:
"```{r}" 
and
"```".

To run code in a chunk, you click on the *Run* button at the top right of the chunk (or you place your cursor inside it and press *Ctrl+Shift+Enter*).

*Exercise*
Run this chunk. Where is the output displayed?
```{r}
5 - 1
```

To run just part of the chunk, you can select the code of interest, r just place your cursor on any line, and press *Ctrl+Enter*.

*Exercise*
Only run the second line of this chunk.
```{r}
# Area of a circle
pi * 4 ^2
pi * 5.7 ^2
```

To add a new chunk, you click on the toolbar: Code -> Insert Chunk.

*Exercise*
Add a new chunk in the space below, and use it to calculate the area of a rectangle of length 3.1 and width 7.5.




# Assignment
You can tell R that something has a name; R will remember it and you can use that later in your code. This is called *assignment* and is done by using the arrow <-.
The following code means: "radius" gets the value of 5.7.

*Exercise*
Run this chunk. In which part of the screen is R storing this value?
```{r}
# Assigning a value to "radius"
radius <- 5.7

# Visualising the value of "radius"
radius
```

From now, you can use radius in the rest of your code, and R will know what you mean.

*Exercise*
Calculate the diameter of a circle of radius 5.7, without typing 5.7.
```{r}

```

# Functions
In programming, a function is like a verb. It takes something (its "arguments"), it does something to it, and gives you back an output.
In R, the arguments of a function are kept in brackets.
```{r}
# Square root function
sqrt(64)
```

*Exercise*
Create a new chunk below, and use it to calculate the square root of 5.7, without typing 5.7.





# Datasets
Some sample datasets are available in R. Let's import "women", a small dataset containing the heights and weights of 15 women, into our Environment.

```{r}
women <- women
```

*Exercise*
Import "chickwts", a dataset containing information on chicken feeds and weights.
```{r}

```


To visualise this dataset, you can run its name, or you can either click on it in the Environment or you can use the function View(). Caution: R is case-sensitive, so you need to type this with an upper-case V, otherwise it won't work.

*Exercise*
Visualise this dataset in one of these two ways. Which one do you prefer?
```{r}
women
View(women)
```

Now let's use the mean() and sd() functions to calculate the average woman height and its standard deviation. R can work with more than one dataset at the time. For this reason, unlike Stata, everytime you refer to a variable, you also need to tell R which dataset is that variable from. You do this by typing a $ sign between the dataset and the variable name.
```{r}
mean(women$height)
sd(women$height)
```

*Exercise*
1. Calculate the `median` weight among `chickens`, and convert it from grams into kilograms
2. Use the formula table() to tabulate the chicken feeds.
```{r}
# Median chicken weight (in kg)


# Chicken feed frequencies


```

*Recap exercise*
Use the women dataset and calculate the mean BMI. There are a few ways of coding this and the choice is yours :)
(BMI = weight in kg divided by squared height in m; 1 lb = 0.453592 kg, 1 in = 0.0254 m)
```{r Solution}
weight <- mean(women$weight) * 0.453592
height <- mean(women$height) * 0.0254
bmi <- weight / height ^2
bmi
```