---
title: "5: Data cleaning"
subtitle: "Introduction to R"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

# What you will learn

* Data cleaning
 - Messy & dirty data
 - Delete variables
 - Rename variables
 - Data classes
 - Label categorical values
 - Recode categorical values
 - Relevel categorical values

-------------------------------------------------------------------------------

```{r}
library("haven")     # import .dta datasets
library("lubridate") # dealing with dates
library("magrittr")  # %$% pipe
library("dplyr")     # data wrangling (part of the tidyverse)
library("ggplot2")   # graphs
library("forcats")   # dealing with factors
```

-------------------------------------------------------------------------------


# Data cleaning

## Exploring messy & dirty data

So far we've worked on the `diab_stata.dta` dataset. Now imagine that data collection has progressed, and now there some new variables.

Unfortunately Bradley, the PI, thought it was a good idea to ask for the data collection to be done in Excel, and he hasn't prepared a data dictionary.

Load the new `diab_new.csv` dataset, along with the old one.
```{r}
# Data import
dm_new <- readr::read_csv("diab_new.csv")

# Old dataset
dm_stata <- read_stata("diab_stata.dta") %>% mutate_if(is.labelled, as_factor)
```
You might have noticed some warnings. Disregard them for now, and try summarising the new dataset with the same functions that we used in the previous sessions with the .dta file. You can also use `View()` to look at the data.

*Exercise 1*
How many major differences can you spot between the two summaries below?

```{r}
# What's happening?
summary(dm_new)

# Compare with 
summary(dm_stata)
```

We'll now work on "cleaning" this new dataset, to make it more useful.

-------------------------------------------------------------------------------

## Deleting variables

It looks like `X18`, `X19` and `test_delete` don't contain any data:
```{r}
dm_new %>% select(c(X18, X19, test_delete))
```
(If you're wondering: "what's `c()`"? See below).

Let's remove these empty variables from the dataset.

You technically assign it a `NULL` value (note that it's uppercase, and different to `NA`, which is a missing value).
A {tidyverse} alternative consists of selecting all variables except these three, and reassigning this as our dataset.
```{r}
#--- Delete variables from the dataset
dm_new$X18 <- NULL
dm_new$X19 <- NULL
```

*Exercise 2*
Delete the last variable.
```{r}

```

```{r}
#--- An alternative
# dm_new %<>% select(-c(X18, X19, test_delete))
```

(Want to learn more? `c()` and vectors)
A "vector" is an ordered list containing objects (of the same data type).
You create a vector with function `c()` – which stands for "concatenate" or "combine".
Vectors are often used when we want to apply a function to more than one object.
```{r}
# log(6)
# log(c(6, 7, 8))
```

-------------------------------------------------------------------------------

## Data classes

One of the main reasons for so many problems with the new dataset is that its *data classes* are not specified.

We know that some variables are categorical, some are ordered, some are quantitative, but `.csv` files do not hold this information. Certain R functions are designed to work with only specific data classes.

R data classes that are relevant to Epidemiology are:

* *factor* (categorical variable)
* *numeric* (continuous variable)
 - integer
 - double (numbers with decimal digits)
* *date*
* *character* (letters)
* *logical* (possible values: `TRUE`, `FALSE`, `NA`)

To check which class an object is, you use `class()`:
```{r}
class(3)
class("Pão de Açúcar")
class(TRUE)
class(dm_stata$gender)
class(dm_new$gender)
```

So, after we import a raw dataset, we need to tell R which classes the variables actually are – and we can decide to give labels to categorical variables, to make our lives easier when we're running the analyses.

In base R, the way we change a variable's class is by running a function on the variable and assigning it to the same variable. For example, to change a variable into a character class:
```{r}
# Check class before
class(dm_new$ID)

# Change class
dm_new$ID <- as.character(dm_new$ID)

#--- Alternative
# dm_new %<>% mutate(ID = as.character(ID))

# Check class after
class(dm_new$ID)
```

-------------------------------------------------------------------------------

### Factors

These are stored as factors. Factors have "levels" indicated by integers (1, 2, 3...) and corresponding "labels" which contain text. For example, in a "outcome" variable, level 1 can represent "alive", level 2 can represent "dead".

Functions `as.factor()` and `factor()` change the class into a factor.

*Exercise 3*
Try summarising the variable `religion` and check its output. Then change its class to factor, and summarise it again.
```{r}

```

We'll now convert all other categorical variables to factor.

```{r}
dm_new$cis_trans <- as.factor(dm_new$cis_trans)
dm_new$sexual_orientation <- as.factor(dm_new$sexual_orientation)
dm_new$continent <- as.factor(dm_new$continent)
dm_new$diet <- as.factor(dm_new$diet)
dm_new$exercise <- as.factor(dm_new$exercise)
```

(Want to learn more? There are faster ways of doing this!)
```{r}
# In just two lines
# cat_variables <- c("gender", "cis_trans", "religion", "sexual_orientation", "continent", "diet", "exercise")
# dm_new[cat_variables] <- lapply(dm_new[cat_variables], as.factor)
```

-------------------------------------------------------------------------------

### Value labels

Value labels are missing in some categorical values because of poor data collection planning: the software that was used in data collection didn't assign labels to numerical values.

You assign labels with function `factor()` by specifying which labels correspond to which levels. Note that any levels that you don't label will be trated as `NA`s.

```{r}
# Check before
table(dm_new$disability, useNA = "ifany")

# Change to categorical and label values
dm_new$disability <- factor(dm_new$disability,
                            levels = c(0, 1, 2),
                            labels = c("able-bodied", "mild", "severe"))

# Check after
table(dm_new$disability, useNA = "ifany")

table(dm_new$HTN, useNA = "ifany")
table(dm_new$CHD, useNA = "ifany")

dm_new$HTN <- factor(dm_new$HTN,
                     levels = c(0, 1),
                     labels = c("no", "yes"))
dm_new$CHD <- factor(dm_new$CHD,
                     levels = c(0, 1),
                     labels = c("no", "yes"))
```

*Exercise 4* 
Now do the same with variable "gender" (1: female, 2: male, 3: non-binary)
```{r}

```

-------------------------------------------------------------------------------

### Dates

Dates are usually imported as character strings; we need to tell R they're, in fact, dates. Remember function `dmy()` from package {lubridate}? It transforms dates stored in any day-month-year format into the correct date class.
```{r}
# Summary before
summary(dm_new$DOB)
class(dm_new$DOB)

# Change to date format
dm_new$DOB <- dmy(dm_new$DOB)

# Summary after
summary(dm_new$DOB)
class(dm_new$DOB)
```
As you can see, after telling R what the correct class is, it can summarise the variable in a much more meaningful way.

-------------------------------------------------------------------------------

## Rename variables

Some variables have unclear, long or uppercase names. Let's rename them so they're easier to understand, or to type.
You can either do this in two ways:
* in base R, by assigning the variable to a new name and then removing the old variable;
* in package {dplyr}, by using function `rename()` and assigning it to our dataset name.
```{r}
# Method 1: base R
dm_new$height <- dm_new$ht
dm_new$ht <- NULL

dm_new$weight <- dm_new$wt
dm_new$wt <- NULL

# Method 2: dplyr
dm_new <- dm_new %>% rename(id = ID,
                            orient = sexual_orientation)

# Method 2 with an assignment pipe
dm_new %<>% rename(htn = HTN,
                   chd = CHD) 

summary(dm_new)
```

*Exercise 5*
Rename variables DOB and BMI as lowercase, with any method you want.
```{r}

```

-------------------------------------------------------------------------------

## Recoding values 

Now let's have a look at variable "diabetes".
```{r}
table(dm_new$diabetes, useNA = "ifany")
```
It looks like there two "0" values were mistyped as "0-". Let's fix this with `recode()`, and let's make it a factor variable.

```{r}
# Change a single value: with dplyr
dm_new %<>% mutate(diabetes = recode(diabetes, "0-" = "0"))

# Method 2 (base R)
# dm_new$diabetes[dm_new$diabetes == "0-"] <- "0"

# Make it categorical
dm_new$diabetes <- factor(dm_new$diabetes,
                          levels = c(0, 1),
                          labels = c("no", "yes"))

# Check it worked
table(dm_new$diabetes, useNA = "ifany")
```

Now let's explore the values of "height" with a Box plot.
```{r}
ggplot(dm_new, aes(height, gender)) + geom_boxplot()
```

It looks like there is an extreme outlier. Let's identify it with function `max()`.
*Exercise 6* Let's correct this value. Note that you need backticks (`) to identify the integer to recode.
```{r}
# Identify the wrong value
max(dm_new$height)

# Correct an integer value

```

-------------------------------------------------------------------------------

## Relevelling factors

Finally, we might sometimes need to reorder the strata in an ordered variable, for example because R has ordered it alphabetically.
We use function `fct_relevel()` from package forcats.
```{r}
# Demonstrate wrong order
ggplot(dm_new, aes(exercise)) + geom_bar()

# Reorder the levels in an ordered variable
dm_new$exercise <- fct_relevel(dm_new$exercise, "none", "sporadic", "regular")

# Check it worked
ggplot(dm_new, aes(exercise)) + geom_bar()
```

-------------------------------------------------------------------------------

*Challenge exercise*
1. Import the `poo.csv` dataset and briefly explore it.
2. Change the data classes to the correct ones.
3. Create a new variable with the age at admission.
4. Cross-tabulate Bristol type with colour, with column percentages.
5. Explore the distribution of temperature according to sex with a Box plot.
6. Save a copy as `poo_cleaned.dta` so you can send it to your colleague Karen, who only uses STATA.

```{r}
# Import and explore


# Change data classes


# Calculate age


# Cross-tab Bristol×Colour


# Temperature Box plot


# Export to Stata


```

-------------------------------------------------------------------------------

## Solutions

*Ex 1 solution*
 * In the new dataset there are 5 new columns: HTN, CHD (which contain meaningful data), X18, X19, test_delete (which do not);
 * There is no meaningful summary for many variables, which now read "class: character".
 * The values of some categorical variables have no labels, only numbers.
 * The variable names are slightly different (capital letters, abbreviations or longer names)

```{r Ex 2 solution}
dm_new$test_delete <- NULL
```

```{r Ex 3 solution}
# Summary before
summary(dm_new$religion)

# Change to factor
dm_new$religion <- as.factor(dm_new$religion)

# Summary after
summary(dm_new$religion)
```

```{r Ex 4 solution}
table(dm_new$gender, useNA = "ifany")

dm_new$gender <- factor(dm_new$gender,
                            levels = c(1, 2, 3),
                            labels = c("female", "male", "non-binary"))
```

```{r Ex 5 solution}
dm_new %<>% rename(dob = DOB, bmi = BMI) 
```

```{r Ex 6 solution}
dm_new %<>% mutate(height = recode(height, `1777` = 177))
```

```{r Challenge exercise solution}
# Import and explore
poo <- readr::read_csv("poo.csv")
poo

# Change data classes
poo %<>% mutate(bed_number = factor(bed_number),
               ward_bay = factor(ward_bay),
               sex = factor(sex),
               date_birth = dmy(date_birth),
               date_admission = dmy(date_admission),
               bristol = factor(bristol),
               colour = factor(colour))

# Calculate age
poo %<>% mutate(age = as.numeric(date_admission - date_birth) / 365.25)

# Cross-tab colour×Bristol
poo %$% summarytools::ctable(bristol, colour, prop = "c")

# Temperature Box plot
poo %>% ggplot(aes(temperature, sex)) + geom_boxplot()

# Export to Stata
poo %>% haven::write_dta("poo_cleaned.dta")
```
