---
title: "3: Summarising data"
subtitle: "Introduction to R"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

# What you will learn

* Pipes
 - Exposition pipe: `%$%`
 - Forward pipe: `%>%`
 
* Data summarisation
 - Categorical variables
 - Continuous variables
 - Grouped summaries with {dplyr}

* Subsetting a dataset with {dplyr}
 - Filtering rows
 - Selecting columns

-------------------------------------------------------------------------------

```{r}
# Install new packages
install.packages("magrittr")
install.packages("summarytools")
# install.packages("epiDisplay") # install this if you have problems with {summarytools}
```

```{r}
# Load packages
library("haven")                 # read .dta files
library("magrittr")              # provides %$% pipe
library("summarytools")          # better tables
library("tidyverse")             # %>% pipe and much more
options(digits = 2, scipen = 10) # reduce signif digits and scientific notation
```

NB: there seems to be an issue with summarytools on Mac computers. If this happens to you, try this solution: https://community.rstudio.com/t/problem-in-installing-summary-tools-package-in-mac/38387. If you still can't work it out, install and load package "epiDisplay" instead, it has similar tabulation functions.

Let's import the diabetes dataset. The `read_dta()` function is part of package "haven".
```{r}
# Import and label
dm <- read_dta("data/diabetes.dta") %>% mutate_if(is.labelled, as_factor)

# Preview the first 10 rows and columns
dm
```

-------------------------------------------------------------------------------


# Pipes

In the chunk above you might have noticed a weird `%>%`. What was that?

"Pipes" are operators that make your code quicker to write and easier to read.
They're *optional*: base R doesn't have them.
We're going to use two pipes: `%$%` and `%>%`.

## %$% pipe

You will remember that `data$variable1` subsets the variable called "variable1" from the dataset "data".
```{r}
dm$diet
```

We use this when a function requires a variable (and not the whole dataset); for example, if we want to get the frequencies, we can use the syntax below.
```{r}
table(dm$diet)
```

The `%$%` pipe, from package magrittr, takes what's on its left (usually, the dataset name) and moves it inside the function, before each argument, and adds a `$` in between. It lets you avoid typing the dataset name and `$` twice, and it makes the true arguments of the function stand out.
It's particularly useful when a function uses multiple variables, because otherwise you would need to specify which dataset is each variable from – this would quickly become verbose.
The two commands below generate the same output.
```{r}
# Without a pipe (= base R)
table(dm$diet, dm$exercise)

# With the %$% pipe
dm %$% table(diet, exercise)
```

*Exercise*
Rewrite the following code using %$%.
```{r ex1}
# Mean height
mean(dm$height)

# With a %$%

```



# %>% pipe

The `%>%` takes what's on the left and it moves it inside the brackets on the right, and follows it with a comma.
It's really helpful to avoid using too many nested functions, or assigning too often. There's a keyboard shorcut for it: *Ctrl + Shift + M* (on Mac: *Cmd + Shift + M*).

These options produce the same output:
```{r}
# Nested functions
log(sqrt(mean(dm$height)))

# Over-assignment
mean_height <- mean(dm$height)
root_mean_ht <- sqrt(mean_height)
log(root_mean_ht)

# Code with %>% pipe
mean(dm$height) %>% sqrt() %>% log()

# Code with two pipes
dm %$% mean(height) %>% sqrt() %>% log()
```

*Exercise*
Rewrite this chunk in just one line by using two pipes. You can break it down into two steps if you prefer.
(`aov()` is a function that performs the statistical test ANOVA, analysis of variance), and takes as argument two variables separated by operator ~).
```{r ex2}
# ANOVA in base R
anova1 <- aov(dm$bmi ~ dm$diet)
summary(anova1)

# ANOVA with two pipes

```


So you can also use `%>%` for any function that takes the dataset as its first argument (like all functions in the tidyverse, for example `ggplot()`.
```{r}
# Without a pipe
ggplot(dm, aes(gender)) + geom_bar()

# With a pipe
dm %>% ggplot(aes(gender)) + geom_bar()
```

-------------------------------------------------------------------------------


# Data summarisation

`summary()` is a useful function that lets you summarise your entire dataset (or just a single variable)

*Exercise*
Run the following chunk.
- What is R doing to each variable?
- What is the median BMI?
- How many missing values does the "exercise" variable have?
```{r}
summary(dm)
```

*NB:* this only works if the variables are coded in the correct data classes, so it will not work well with csv datasets without cleaning them first. See `intRo5 clean.Rmd` for how to do this.

## Summarising categorical variables: frequency tables

To tabulate categorical data, we have many options, among which `table()` from base R, `count()` from {dplyr}, `freq()` from package {summarytools}, `tab1()` from {epiDisplay}.

*Exercise*
Run this chunk and compare the output from those three functions.
- What's the difference?
- Which function do you prefer?
- What does the `useNA` argument do? (Try removing it, run the code again, and see what happens)
```{r}
# Simple frequency table
table(dm$gender)
count(dm, gender)
freq(dm$gender)
epiDisplay::tab1(dm$gender, graph = F)
```

```{r}
# More complex frequency tables
table(dm$sexual_orien, useNA = "ifany")
freq(dm$sexual_orien, report.nas = TRUE, cumul = TRUE)
```

*Exercise*
What percentage of all people in this sample has a mild disability?
```{r ex freq}

```


## Summarising categorical variables: cross-tabulation

To make two-way tables you can use
* `table()` from base R, but it doesn't calculate percentages for you.
* `count()` from {dplyr}, but it's in nested rows and doesn't give you percentages.
* `ctable()` from {summarytools}; it's a very customisable function. 
* `tabpct()` from {epiDisplay}.

```{r}
# Base R
dm %$% table(diet, diabetes, useNA = "ifany")

# Package dplyr
dm %>% count(diet, diabetes)

# Package summarytools
dm %$% ctable(diet, diabetes, headings = FALSE, prop = "c", useNA = "no")

# Package epiDisplay
dm %$% epiDisplay::tabpct(diet, diabetes, graph = F, percent = "col")
```

You will note that different functions have slightly different ways of doing some things:
- including or omitting missing values (e.g. useNA in table() and ctable())
- deciding whether to show column or row percentages/proportions (e.g. prop = "c" or "r" in ctable(), percent = "col" or "row" in tabpct()). 


*Exercise*
What percentage of trans people is atheist?
Hint: if you can't remember the names of your variables, you can use function `names()`, which takes the dataset name as its argument.
```{r ex contingency}

```

## Summarising continuous variables

As we've seen previously, to summarise continuous variables, you can use functions `mean()`, `median()`, `sd()` – or you can use `summary()` again:
```{r}
mean(dm$weight)
median(dm$height)
sd(dm$bmi)
summary(dm$exercise_weekly)
```

# Grouped summaries with {dplyr}

What if we wanted to calculate the mean weight according to continent?

We can use two powerful functions from package {dplyr}, part of the {tidyverse}: `group_by()` and `summarise()`. Combined, these two functions collapse the dataset according to a categorical variable, and summarise it with on or more other functions.

```{r}
dm %>% group_by(continent) %>% 
       summarise(mean(weight))
```

We can also summarise more than one variable:
```{r}
dm %>% group_by(diabetes) %>% 
       summarise(mean(height),
                 median_exercise = median(exercise_weekly, na.rm = T))
```

And we can group by more than one variable. In these cases it's useful to have a count of individuals in that combination, with `n()`.
```{r}
dm %>% group_by(diabetes, diet) %>% 
       summarise(count = n(),
                 median_BMI = median(bmi))
```

You can summarise with any of these functions:
- `mean()`, `median()`,
- `sd()`, `IQR()`,
- `min()`, `max()`,
- `n()`, `n_distinct()`

-------------------------------------------------------------------------------

# Transforming data with dplyr

## Filter rows

Sometimes you don't need to use all the observations (rows) in your dataset, but only a group of those: for example, you might want to check the median height only in women. The function `filter()` lets you do just that: it filters your dataset according to a condition. You can then pipe this into any other function.

```{r}
# Only keep females
dm %>% filter(gender == "female")
# note that it's a double = sign!

# Pipe this filtered dataset into the median() function
dm %>% filter(gender == "female") %$% median(height)
```

You can also filter on more than one condition; for example, you might want to only filter underweight women.
```{r}
dm %>% filter(gender == "female", bmi < 18.5)
```

*Exercise*
Filter Christian men shorter than 170 cm with a mild disability. How many are in our dataset?
Hint: if you need to check which values are possible in a categorical variable, you can use `levels()`, which takes the variable as an argument.
```{r filter}

```




`filter()` can be very useful to make graphs with only a subgroup of data. Compare these two pieces of code:
```{r}
# Plotting all observations
dm %>% ggplot(aes(exercise_weekly, bmi)) +
  geom_point(aes(shape = diabetes, colour = gender)) +
  geom_smooth(method = "lm")

# Filtering out inactive people
dm %>% filter(exercise_weekly > 0) %>%
  ggplot(aes(exercise_weekly, bmi)) +
  geom_point(aes(shape = diabetes, colour = gender)) +
  geom_smooth(method = "lm")
```

*Exercise*
Create the same scatterplot, but without men.
```{r scatter}
# Base scatterplot
dm %>% ggplot(aes(height, weight)) + geom_point(aes(colour = gender)) + ylim(c(25,125)) + xlim(c(130,200))

# Scatteplot without men

```


## Select columns

Function `select()` selects only certain columns (variables). Imagine we wanted a dataset containing only information about ID and height:

```{r}
dm %>% select(id, height)
```

We can add as many variables as we need.

*Exercise*
Select only variables with nutritional information.
```{r select}

```

You can combine `select()` and `filter()` by piping. For example, you might want to only keep information on exercise and diabetes for Hindu men:
```{r}
dm %>% filter(religion == "hindu", gender == "male") %>% 
  select(id, exercise, exercise_weekly, diabetes)
```

Note that `filter()` and `select()`, like the other {dplyr} functions, _do not_ change the original dataset permanently. If you want to keep a copy of a transformed dataset, you need to assign it a name.
```{r}
dm_overweight <- dm %>% filter(bmi >= 25 & bmi < 30) %>%
                        select(id, gender, bmi, diabetes)
```

*Recap exercise*
1. Create a new dataset called "dm_tall" containing only data on ID, gender, cis/trans, sexual orientation, and religion, only for able-bodied people who are 1.95 metres tall or above.
2. How many people satisfy these filters?
3. What are their IDs?
```{r recap}

```

-------------------------------------------------------------------------------

# Solutions

```{r ex2 solution}
dm %$% aov(height ~ weight) %>% summary()
```

```{r ex freq solution}
freq(dm$disability, report.nas = TRUE, cumul = F)
```

```{r ex contingency solution}
dm %$% ctable(religion, cistrans, total = FALSE, headings = FALSE, prop = "c")
```

```{r filter solution}
dm %>% filter(gender == "male", height < 170, disability == "mild")

```

```{r scatter solution}
dm %>% filter(gender != "male") %>% ggplot(aes(height, weight)) + geom_point(aes(colour = gender)) + ylim(c(25, 125)) + xlim(c(130,200))
```

```{r select solution}
dm %>% select(id, weight, bmi, diet, diabetes)
```

```{r recap solution}
dm_tall <- dm %>%
  filter(disability == "able-bodied", height > 195) %>%
  select(id, gender, cistrans, sexual_orien, religion)

dm_tall
```

-------------------------------------------------------------------------------
