---
title: "9: Analysing binary outcomes"
subtitle: "Introduction to R for health data"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
---

NB: work in progress

------------------------------------------------------------------------

# What you will learn

-   Introduction to data analysis of binary outcomes
    -   chi-square test
    -   logistic regression


------------------------------------------------------------------------

# Recap from topic 8

...

------------------------------------------------------------------------

```{r}
library(magrittr)
library(summarytools)
library(tidyverse)
library(gtsummary)
library(broom)
```

------------------------------------------------------------------------

# Analysing binary outcomes

```{r}
viral <- rio::import("data/viral.rds")
viral
```


Visualise and summarise before you infer

```{r}
viral |>
  ggplot(aes(y = ms, fill = ebv_baseline)) +
  geom_bar(position = "dodge") +
  coord_fixed(ratio = 120)
```

```{r}
viral %$% ctable(ebv_baseline, ms)
```



## Chi-square test

```{r}
# Base R
chisq_ms_ebv <- viral %$%
  table(ms, ebv_baseline) |>
  chisq.test()
chisq_ms_ebv
chisq_ms_ebv$expected |> round(1)


# Using summarytools
viral %$% ctable(ms, ebv_baseline, prop = "c", chisq = TRUE, OR = TRUE)
```

*Exercise 1.*

1. Visualise the relationship between sex and MS.
2. Summarise this with a contingency table.
3. Perform a chi-square test and calculate the OR.





------------------------------------------------------------------------

## Logistic regression

### With a binary exposure

```{r}
logistic_ebv <- glm(ms ~ ebv_baseline,
                    family = binomial,
                    data = viral)

summary(logistic_ebv)

tbl_regression(logistic_ebv, exponentiate = TRUE)
```


### With a continuous exposure

```{r}
logistic_age <- glm(ms ~ age_baseline,
                    family = binomial,
                    data = viral)

summary(logistic_age)

tbl_regression(logistic_age, exponentiate = TRUE)
```


### Multivariable


```{r}
logistic_multi <- glm(ms ~ ebv_baseline + age_baseline,
                       family = binomial,
                       data = viral)

summary(logistic_multi)

tbl_regression(logistic_multi, exponentiate = TRUE)

# Compare with previous
tbl_regression(logistic_ebv, exponentiate = TRUE)
```

------------------------------------------------------------------------


More resources

-  [Trap of trends to statistical significance, BMJ 2014](https://www.bmj.com/content/348/bmj.g2215)
Solutions
- Infer package - for tidy tests including `t_test()` and `chisq_test()` wrappers
- broom package - to improve output of regression models

```{r ex1_solution}
# Q1
viral |>
  ggplot(aes(y = ms, fill = sex)) +
  geom_bar(position = "dodge")

# Q2 and Q3
viral %$% ctable(sex, ms, chisq = TRUE, OR = TRUE)
```