---
title: "6: Analysing continuous outcomes"
subtitle: "Introduction to R"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
---

-------------------------------------------------------------------------------

# What you will learn

* Data analysis of continuous outcomes
 - data visualisation
 - t-test
 - linear regression

-------------------------------------------------------------------------------

```{r}
library(magrittr)
library(tidyverse)
options(scipen = 99, digits = 2)
```

-------------------------------------------------------------------------------


## Introduction

Let's load the Ambition dataset. This contains some simulated result of the AMBITION study, a clinical trial investigating the use of single-dose liposomal amphotericin B (LAmB) to treat cryptococcal meningitis, instead of the standard (longer) course.

LAmB is nephrotoxic, so we're interested in the renal function in these two groups.
```{r}
amb <- readRDS("data/ambition.rds")
```

Just to familiarise yourself with the dataset:

* Exposure variable: "arm" (single-dose LAmB vs standard LAmB course)

* Outcome variable: "crea_2w" (serum creatinine level at 2 weeks)

* Baseline characteristics:
 - "age" (integer)
 - "sex" (binary)
 - "country" (categorical)
 - "crea_0" (baseline creatinine, in mmol/L)
 - "egfr_0" (baseline eGFR calculated with MDRD equation)

-------------------------------------------------------------------------------

## Average outcome value by a binary exposure

Let's imagine our research question is: "After starting LAmB, is serum creatinine level at 2 weeks different depending on the dose?"

To have a first idea we can summarise the creatinine outcome in the two treatment arms, by calculating an average.

```{r}
amb %>% group_by(arm) %>% summarise("number of patients"= n(),
                                    "mean creatinine at 2w" = mean(crea_2w))
```
The average creatinine at 2 weeks is 110 mmol/L in the control group, and 98 mmol/L  in the single-dose group. In this sample, there's a difference of 12 mmol/L. How do we know if this reflects a true difference in the whole population, and is not due to chance?

-------------------------------------------------------------------------------

## Visualising a continuous outcome by a binary exposure

Let's get more insight by visualising the data.

We have a categorical exposure variable (treatment arm) and a continuous outcome variable (creatinine at 2 weeks). Which of these graphs are more appropriate?
- bar chart
- Box plot
- histogram
- scatter plot
- violin plot

*Exercise 1*
Use an appropriate graph to visualise the distribution of creatinine at 2 weeks according to treatment arm.
```{r}

```

From your data visualisation, you can see that there is quite a lot of overlap between creatinine values in both groups, however a large portion of the interquartile ranges don't overlap.

This is helpful to understand the data better, but still doesn't tell us if this difference could be due to chance.

-------------------------------------------------------------------------------

## Confidence intervals around a mean

We can calculate the confidence intervals around each mean.

The formula for the confidence interval of a mean depends on the mean itself and the standard error (which itself depends on the standard deviation and the sample size). For a 95% CI, we multiply the SE by 1.96.

```{r}
amb %>% group_by(arm) %>% summarise(
                      "number of patients"= n(),
                      "mean creatinine at 2w" = mean(crea_2w),
                      "lower CI" = mean(crea_2w) - 1.96 * sd(crea_2w) / sqrt(n()),
                      "upper CI" = mean(crea_2w) + 1.96 * sd(crea_2w) / sqrt(n())
)
```

An intuitive interpretation of the second confidence interval is:
"We can be 95% confident that the mean true population mean of creatinine at 2 weeks among patients treated with a single LAmB dose lies between 96 and 101".

As you can see, with a continuous outcome even relatively small sample sizes give us very precise statistical estimates.

-------------------------------------------------------------------------------

## t test

In order to do hypothesis testing on the difference between two means, we can use a t test (unpaired because this data is not paired). 

The function for the t test is `t.test()`.

In R statistical tests, variables listed before the tilde `~` are the outcome/response/dependent variables, and variables after the tilde are the exposure/independent variables.

Our hypothesis is that creatinine at 2 weeks is different according to how much LAmB did the patients receive.
Our alternative hypothesis is that there is no difference (difference in means = 0).

```{r}
amb %$% t.test(crea_2w ~ arm)
```

The key output here is the p-value, <0.001. This means that there is very strong evidence against the null hypothesis, and so very strong evidence that creatinine at 2 weeks is indeed different according to treatment group.

The other important output is the 95% CI of this difference: (8.3, 15.0). We can be 95% confident that the true difference in creatinine average is between 8.3 and 15.0 mmol/L.

*Exercise 2*
Now try assessing with a t-test whether *baseline* creatinine is linked to treatment group.

```{r}

```
How do you interpret these results?


*Challenge exercise 3*
Assess whether baseline eGFR is different in people from Uganda compared to people from Malawi.

Hint: the "country" variable contains more than 2 countries. How do restrict the dataset only to people from Uganda and Malawi?
```{r}

```

(What about if you have more than 2 groups in your exposure variable? You can't use a t test, you need to use ANOVA.)

-------------------------------------------------------------------------------

## A continuous outcome by a continuous exposure

Now let's go through another research question: "Is creatinine at baseline related to age?"

Both variables are continuous now, so we can't calculate mean creatinine according to age unless we categorised the age group. What we can do is plotting this.

*Exercise 4*
What's the best way to plot two continuous variables against each other?
- bar chart
- Box plot
- histogram
- scatter plot
- violin plot

Plot creatinine at baseline against age. Do you think there's a relationship?

```{r}

```

-------------------------------------------------------------------------------

## Linear regression

Linear regression is used to conduct hypothesis testing with a continuous outcome and one or more exposure variables. If there's just one binary exposure variable, it will be equivalent to a t-test.

What's great about linear regression is that we can visually represent it. In {ggplot2}, we can do this with `geom_smooth()`:

```{r}
amb %>% ggplot(aes(x = age, y = crea_0)) +
  geom_point() +
  geom_smooth(method = "lm") # lm stands for linear model
```

What do you think of this regression line? Is there a relationship between creatinine and age?

The null hypothesis in linear regression is that in the population, this regression line is flat: in other words, there is no relationship between our exposure (age) and our outcome (creatinine). This is numerically equivalent to saying that the slope of that line is 0.

The alternative hypothesis is that the slope is somewhat different to 0.

The function is `lm()` (linear model) and annoyingly it follows a different syntax to `t.test()`. The dataset is specified as the "data" argument. We can then pipe into the `summary()` function to get the output we need.
```{r}
lm(crea_0 ~ age,
   data = amb) %>%
  summary()
```

We're interested in these key outputs:
- The intercept estimate (In this case, 55 mmol/L)
- the age coefficient estimate (In this case, 0.53)
- The p-value for the age estimate - visualised as Pr(>|t|) - in this case, <0.001.

From the intercept and the coefficient, we can write a regression equation:

creatinine = 0.53 * age + 55

This means that for every year of life, creatinine increases by 0.53 mmol/L.

How confident are we about this statement? Very confident: the p value is <0.001. We can say that there is very strong evidence against the null hypothesis.

(The intercept, 55, mathematically refers to the mean creatinine at age 0 years. But in this dataset the youngest person is 18, so that value is purely extrapolated by continuing the regression line to the left. It does not have any clinical meaning.)

*Exercise 5*
Now assess the relation between eGFR at baseline and age:
1. Draw a scatterplot with a linear regression line. What do you think of the distribution?
2. Conduct hypothesis testing with a linear regression model.
3. Write the regression equation. What does this mean?
4. How confident are you that this results reflects the true population values?
```{r}
# 1: scatterplot


# 2: linear model


# 3: regression equation


# 4: interpretation


```

-------------------------------------------------------------------------------

*Challenge exercise 6*
Creatinine at 2 weeks is not very meaningful - in real life you'd compare that value with the baseline creatinine, and this is what the trial is looking at.
1. Create a new variable that represents the percent change in creatinine.
2. Visualise the distribution of this change in the two treatment arms.
3. Use an appropriate statistical test to assess whether there's a relation between change in creatinine and treatment arm.
```{r}
# 1


# 2


# 3


```




---

Solutions

```{r Exercise 1}
# Option A: Box plot
ggplot(amb, aes(y = crea_2w, x = arm)) +
  geom_boxplot() +
  coord_fixed(ratio = 0.02)

# Option B: violin plot
ggplot(amb, aes(y = crea_2w, x = arm)) +
  geom_violin(draw_quantiles = 0.5) +
  coord_fixed(ratio = 0.02)

# Option C: stratified histogram
ggplot(amb, aes(x = crea_2w, fill = arm)) +
  geom_histogram(binwidth = 12.5) +
  facet_grid(rows = vars(arm))
```


```{r Exercise 2}
amb %$% t.test(crea_0 ~ arm)
```

```{r Challenge exercise 3}
amb %>%
  filter(country == "uganda" | country == "malawi") %$%
  t.test(egfr_0 ~ country)
```

```{r Exercise 4}
amb %>% ggplot(aes(x = age, y = crea_0)) + geom_point()
```

```{r Exercise 5}
# 1: scatterplot
amb %>% ggplot(aes(x = age, y = egfr_0)) +
  geom_point(colour = "darkgreen") +
  geom_smooth(method = "lm", colour = "purple")

# 2: linear model
lm(egfr_0 ~ age,
   data = amb) %>% 
  summary()

# 3: regression equation
# eGFR = -1.34 * age + 129

# 4: interpretation
# p-value < 0.001. There's very strong evidence that eGFR decreases with age.
```

```{r Challenge exercise 6}
# 1
amb %<>% mutate(change_creat = 100*(crea_2w - crea_0)/crea_0)
amb %>% select(ID, crea_0, crea_2w, change_creat)

# 2
amb %>% ggplot(aes(arm, change_creat)) + geom_boxplot()

# 3, option A
amb %$% t.test(change_creat ~ arm)

# 3, option B
lm(change_creat ~ arm,
   data = amb) %>% 
  summary()
```

