---
title: "7: Reshaping data"
subtitle: "Introduction to R for health data"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
---

NB: work in progress

-------------------------------------------------------------------------------

# What you will learn

* Join multiple datasets
- Append rows
- Add columns: left join

* Reshape a dataset
- Pivot longer
- Pivot wider


-------------------------------------------------------------------------------

```{r}
library(readr)
library(dplyr)
library(tidyr)
```


TODO: decide which dataset you want to use.

```{r}
dat <- tribble(
    ~id, ~age, ~pcr, ~hb_1, ~hb_2,
        1, 65, "neg", "neg", "neg",
        2, 21, "pos", "neg", "neg",
        3, 57, "pos", "pos", "neg",
        4, 84, "neg", "neg", "pos",
        5, 75, "neg", "pos", "neg"
        )

dat2 <- tibble(
  id = 1:5,
  gender = sample(c("f", "m"), replace = T, size = 5),
  age = c(63, 81, 73, 75, 86),
  creat = c(55, 47, 54, 77, 241)
) |> 
  mutate(
    creat_2 = creat + rnorm(n = 1, mean = 12, sd = 4) |> round()
  )

dat2


data.frame(
  patient_id = c("a", "a", "b", "b", "b"),
  species = c("C. albicans", "C. glabrata", "C. albicans", "C. krusei", "C. parapsilosis"),
  date_sampling = lubridate::dmy(c("01/05/2022", "06/03/2022", "14/05/2022", "18/08/2022", "23/11/2022")),
  mic_1 = c("S", "R", "S", "R", "I"),
  mic_2 = c("S", "R", "I", "R", "R")
)


```

```{r}
antib_A <- dat |> select(-IgM) |> filter(id < 4)
ren <- dat2 |> select(-creat_2) |> filter(id < 4)
ren
```

```{r}
antib_B <- dat |> select(-IgM) |> filter(id > 3)
ren_more_pts <- dat2 |> select(-creat_2) |> filter(id >3 )
ren_more_pts
```

```{r}
antib_2 <- dat |> select(id, IgM) |> filter(id < 4)
ren_yr2 <- dat2 |> select(id, creat_2) |> filter(id != 5) |> bind_rows(tibble(id = 6, creat_2 = 21))
ren_yr2
```


-------------------------------------------------------------------------------

# Joining data

Often we don't get all the data in a single dataset - we need to join data from different files.

This can be either because there are *new observations* (rows), or because there are *new variables* (columns), or both.


-------------------------------------------------------------------------------

## Joining extra rows

Suppose we have some initial data on serum creatinine levels in a group of outpatients.
```{r}
# TODO: add read_csv(...)
ren
```

The PI decides to extend recruitment and you're sent a new dataset that only contains the new observations. Note that there are overlapping columns.

```{r}
ren_more_pts
```

We now need to combine these two datasets into one.

`bind_rows()` from package `dplyr` can do that. It takes as arguments the two datasets.

```{r}
ren_all_pts <- bind_rows(ren, ren_more_pts)
ren_all_pts
```

This function works even if there is not complete column overlap - but in that case some missing values will be generated.


-------------------------------------------------------------------------------

## Join extra columns

Now imagine that these patients have been followed up for a year and their creatinine levels were measured again.

You have this data in a new dataset:
```{r}
ren_yr2
```

Note that:
1. Both datasets share a unique `id` variable;
2. the patient with id 5 is not present in this dataset (they did not attend the follow-up appointment), so another participant was recruited (id 6) and their demographics were recorded.

We now need to decide how to join these two datasets.

Options:
1. Only include participants from the original dataset, with their updated creatinine measurements. `left_join()`.



```{r}
left_join(ren_all_pts, ren_yr2, by = "id")
```



```{r}
right_join(ren_all_pts, ren_yr2, by = "id")
```



```{r}
full_join(ren_all_pts, ren_yr2, by = "id")
```




```{r}
inner_join(ren_all_pts, ren_yr2, by = "id")
```



-------------------------------------------------------------------------------

# Reshape data


-------------------------------------------------------------------------------

## Pivot longer


-------------------------------------------------------------------------------

## Pivot wider



-------------------------------------------------------------------------------





