---
title: "1: The basics"
subtitle: "Introduction to R"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

# What you will learn

* User interface
* Console
* R Markdown notebooks
* Assignment
* Functions
* Datasets

-------------------------------------------------------------------------------

# User interface

The RStudio interface is divided into four parts:
- *top left*: the code script (*.R) or notebook (like this one, intRo1.Rmd) you're working on.
- *bottom left*: the console (where you can input your commands)
- *top right*: Environment (where your datasets and other "objects" are stored)
- *bottom right*: help (and other stuff)

-------------------------------------------------------------------------------

# Console

The console is used to run simple commands that you don't need to save.

*Exercise*
In the console below, add 3 and 5. You do this by using it as a calculator: 3+5 Enter.

Now forget about the console forever – it's better to keep your code in a notebook (or in a script), like in Stata you would with a .do file.

-------------------------------------------------------------------------------

# R Notebooks (.Rmd)

A notebook, like this one (`intRo1 basics.Rmd`) is composed of two things:

* Text, which is the main component
* "Chunks", which contain code (and comments, which are preceded by #)

A chunk looks like this:
```{r}

```
Please note that those three signs are not normal apostrophes ('): they are "backticks" (`). On a Windows keyboard, the key is to the left of the number 1.

To run code in a chunk, you click on the *Run* (green triangle) button at the top right of the chunk (or you place your cursor inside it and press *Ctrl+Shift+Enter* (on Mac: *Cmd+Shift+Enter*).

*Exercise*
Run this chunk. Where is the output displayed?
```{r}
5 - 1
```

To run just part of the chunk, you can select the code of interest (or just place your cursor on any line) and press *Ctrl+Enter* (on Mac: *Cmd+Enter*).

*Exercise*
Only run the second line of this chunk.
```{r}
# Area of a circle
pi * 4 ^2
pi * 5.7 ^2
```

To add a new chunk, you click on the toolbar: Code -> Insert Chunk.

*Exercise*
Add a new chunk in the space below, and use it to calculate the area of a rectangle of length 3.1 cm and width 7.5 cm (multiply).




-------------------------------------------------------------------------------

# R scripts (.R)

An R script, for example `analysis.R`, mainly contains code, and any non-code text needs to be highlighted with a preceding `#`. You can think of R scripts as a very large chunk.

We won't be using R scripts in this course, but you might see them elsewhere. 

-------------------------------------------------------------------------------

# Assignment

You can tell R that something has a name; R will remember it and you can use that later in your code. This is called *assignment* and is done by using the arrow <-. Assignment creates a new "object".
The following code means: "radius" gets the value of 5.7. "radius" is a new object.

*Exercise*
Run this chunk. In which part of the screen is R storing this value?
```{r}
# Assigning a value to "radius"
radius <- 5.7

# Visualising the value of "radius"
radius
```

From now, you can use the object "radius" in the rest of your code, and R will know what you mean.

*Exercise*
Calculate the circumference of a circle of radius 5.7, without typing 5.7 (Formula: twice the radius times pi)
```{r}

```

Note that assigning a new value to an existing object will permanently replace the old version.
```{r}
radius <- 6.58
radius
```


-------------------------------------------------------------------------------

# Functions

In programming, a function is like a verb. It takes something (its "arguments"), it does something to it, and gives you back an output.
In R, the arguments of a function are written in brackets.
```{r}
# Square root function
sqrt(64)
```

*Exercise*
Create a new chunk below, and use it to calculate the square root of 5.7, without typing 5.7.



-------------------------------------------------------------------------------

# Datasets

Some sample datasets are available in R. Let's import "Theoph", a small dataset regarding the pharmacokinetics of theophylline, into our Environment. (It's a default R dataset, so R already knows what it is). We will call it "data_theo" but we could call it whatever name we like.

```{r}
data_theo <- Theoph
```

*Exercise*
Import "chickwts", another default dataset containing information on chicken feeds and weights. You can assign it to whatever name you prefer.
```{r}

```


To visualise this dataset, you can run its name, or you can either click on it in the Environment or you can use the function View(). Caution: R is case-sensitive, so you need to type this with an upper-case V, otherwise it won't work.

*Exercise*
Visualise this dataset in one of these two ways. Which one do you prefer?
```{r}
data_theo
View(data_theo)
```

Now let's use the `mean()` and `sd()` functions to calculate the average theophylline concentration and its standard deviation. That variable is called "conc" in this dataset.
R can work with more than one dataset at the time. For this reason (unlike Stata) every time you refer to a variable, you also need to tell R which dataset is that variable from. You do this by typing a `$` sign between the dataset and the variable name.
(The $ sign is technically called the "exposition operator")
```{r}
mean(data_theo$conc)
sd(data_theo$conc)
```

*Exercise*
1. Calculate the `median` chicken weight, and convert it from grams into kilograms
2. Use the function `table()` to tabulate the types of chicken feeds. This function can use the same syntax as the functions above.
```{r}
# Median chicken weight (in kg)


# Chicken feed frequencies


```

*Recap exercise*
Use the theophylline dataset and calculate the mean subject weight, and convert it into pounds. There are a few ways of coding this and the choice is yours :)
(1 kg = 2.20462 pounds)

-------------------------------------------------------------------------------

# Solutions

```{r}
# Second-to-last exercise
median(chickwts$weight) * 0.001
table(chickwts$feed)

# Recap exercise

# Multiple ways of doing this calculation
# Option 1: all in one line
mean(data_theo$Wt) * 2.20462

# Option 2: break it down
mean_kg <- mean(data_theo$Wt) # calculate the mean weight in kilograms and assign this

mean_lb <- mean_kg * 2.20462   # convert into pounds and assign this

mean_lb                       # visualise result
```

-------------------------------------------------------------------------------
