---
title: "7: Reshaping data"
subtitle: "Introduction to R for health data"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
---

------------------------------------------------------------------------

# What you will learn

-   Join multiple datasets
    -   Append rows
    -   Join columns
-   Reshape a dataset
    -   Pivot longer
    -   Pivot wider

------------------------------------------------------------------------

# Recap from topic 6

1. Run this chunk to create a dirty dataframe containing information on patients with myocardial infarctions seen by different ambulance teams.
2. Clean the column names.
3. Clean the third column so that you calculate mean and standard deviation.
4. Replace "NW" with "North West" in column 2.

```{r}
ambulances <- data.frame(
  County = c("Derby", "Bedford", "Norfolk", "Merseyside", "Cheshire", "Hampshire", "Kent", "Plymouth", "Swindon", "Wiltshire"),
  Region = c("East Midlands", "East of England", "East of England", "NW", "North West", "South East", "South East", "South West", "South West", "South West"),
  Patients.with.a.pre.hospital.diagnosis.of.suspected.STEMI.confirmed.on.ECG = c("104", "105", "266", "140", "188", "No data", "236", "78", "190", "70"))

ambulances
```

------------------------------------------------------------------------

```{r}
library(rio)
library(dplyr)
library(tidyr)
library(ggplot2)
```

------------------------------------------------------------------------

# Joining data

Often we don't get all the data in a single dataset - we need to join data from different tables.

This can be either because there are *new observations* (rows), or because there are *new variables* (columns), or both.

Throughout this topic we'll use data from a study of candidaemia. All data are in a single Excel file, `candidaemias.xlsx`, but in different sheets.

Please open the file in Excel first (you can find it and open it directly via the RStudio user interface: bottom right, Files.)

```{r}
sheets <- c("site5", "site7", "site8", "cfupcr", "demographics", "meds", "outcomes")

for (sheet in sheets) {
  assign(sheet, import("data/candidaemias.xlsx", sheet = sheet))
}
```

------------------------------------------------------------------------

## Append rows

It looks like the β-D-glucan (`bdg`) results are scattered over three separate sheets, one per study site.

```{r}
site5
site7
```

However, they contain the same type of information, so it makes sense to join them in a single dataframe. We could do this in Excel, but that would be time-consuming, boring and error-prone if we had dozens of sheets, each containing thousands of rows.

`bind_rows()` from package `dplyr` lets you bind two dataframes by rows. You can then assign this combined dataframe to a new object.

```{r}
lab <- bind_rows(site5, site7)

lab
```

*Exercise 1.*

1.  Try to add the data from site 8 to this new combined dataset.
2.  What happens?
3.  How can you fix this?

------------------------------------------------------------------------

## Join columns

Now we want to join lab results with outcome data. They're in two separate dataframes:

```{r}
lab
outcomes
```

The join functions from `dplyr` add columns from one dataset to another, using a key to match rows. If the two dataframes don't contain the same rows, we need to specify which rows to include - by picking a specific join function.

Note that, in this case:

1.  Both datasets share a unique `id` variable;
2.  `lab` contains information on many patients that are not listed in `outcomes` - because `outcomes` only lists patients who died.
3.  Similarly, `outcomes` contains data on patients who are not listed in `lab` - because not everyone who died had a candidaemia.

We now need to decide how to join these two datasets.

| Aim                                                      | dplyr function |
|----------------------------------------------------------|----------------|
| Only include participants that are present in `lab`      | `left_join()`  |
| Only include participants that are present in `outcomes` | `right_join()` |
| Only include participants that are present in both       | `inner_join()` |
| Include all participants                                 | `full_join()`  |

Compare what happens:

```{r}
left_join(lab, outcomes, by = "id")
right_join(lab, outcomes, by = "id")
inner_join(lab, outcomes, by = "id")
full_join(lab, outcomes, by = "id")
```

*Exercise 2.*

1.  Join demographic data to the dataframe containing lab results.
2.  (Extra practice) Make a scatterplot to explore the relation between age and BDG at baseline, with different species identified by different shapes.

*Exercise 3.*

Run the chunk below which joins data from `meds` and `lab`.

1.  Do you notice anything odd?
2.  Why do you think this happened?

```{r}
left_join(lab, meds, by = "id")
```

------------------------------------------------------------------------

## Complex joins

Some datasets are more complex. For example, there can be many variables that are repeated in multiple columns, like in this case, timing, CFU count and PCR result in samples A-C.

```{r}
cfupcr
```

Reshaping this to a long format is more complex; we need to use other arguments of `pivot_wider()`, for example `names_sep =` or `names_pattern =`

```{r}
cfupcr |> 
  pivot_longer(!id,
               names_to = c(".value", "sample"),
               names_sep = "_")
```

To learn more about complex joins, please see the Resources section further below.

------------------------------------------------------------------------

# Reshaping data

Let's look again at the `lab` dataframe. Imagine we wanted to make a line graph, to show the BDG through time.

Unfortunately, it the current format, we can't, because the BDG values are scattered across seven columns, and the value for the time points are hidden in the column names.

The best we could do is something like this:

```{r fig.height=5}
lab |> ggplot(aes(shape = species, colour = factor(id))) +
  geom_point(aes(x = 0, y = bdg_0)) +
  geom_point(aes(x = 12, y = bdg_12)) +
  geom_point(aes(x = 24, y = bdg_24)) +
  geom_point(aes(x = 36, y = bdg_36)) +
  geom_point(aes(x = 48, y = bdg_48)) +
  geom_point(aes(x = 72, y = bdg_72)) +
  geom_point(aes(x = 168, y = bdg_168)) +
  scale_x_continuous(breaks = seq(0, 168, 24))
```

This happens because data was collected in a "wide" format: repeated observations for the same patient are recorded in different columns.

This helps with data collection, but it make many types of data manipulation and analysis more complicated.

------------------------------------------------------------------------

## Pivot longer

We can turn this into a "long" format dataframe by using `pivot_longer()`, from `tidyr`.

At the bare minimum, we need to specify which columns need to be reshaped.

I also recommend using the arguments `names_to =` and `values_to` to rename the two new columns that will contain the old column names and the old values (if you don't, they will get the nondescript default names: "name" and "value")

```{r}
lab |> pivot_longer(cols = starts_with("bdg"),
                    names_to = "hours",
                    values_to = "bdg")
```

NB: this has the side effect of duplicating IDs and species! It's an unavoidable evil, but you need to be aware of it.

In order to turn the `hours` column into a numeric class we need to first remove the now unneeded "bdg\_" prefix - we can do this by adding a `names_prefix =` argument:

```{r}
lab_long <- lab |>
  pivot_longer(starts_with("bdg"),
               names_to = "hours",
               names_prefix = "bdg_",
               values_to = "bdg") |>
  mutate(hours = as.numeric(hours))

lab_long
```

```{r fig.height=5}
lab_long |>
  ggplot(aes(x = hours, y = bdg)) +
  geom_point(aes(shape = species, colour = factor(id))) +
  geom_line(aes(linetype = species, colour = factor(id))) +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 12, 24, 36, 48, 72, 168)) +
  theme(panel.grid.minor.x = element_blank())
```

Let's now look at a WHO dataset of new cases of tuberculosis among young women in South American countries in 2010.

```{r}
# Just run this chunk to generate the data
# No need to understand this code if you're short in time

s_america <- tibble(
  country = c("Brazil", "Colombia", "Bolivia (Plurinational State of)",
              "Argentina", "Venezuela (Bolivarian Republic of)", "Uruguay",
              "Paraguay", "Ecuador", "Peru", "Suriname", "Guyana", "Chile"),
  popul = c(216182201, 52571220, 12039798, 46148799, 27360459,
            3501838, 7335451, 18331909, 34234477, 598919, 794910, 19682578))

tb_amer <- who |>
  filter(year == 2010, country %in% s_america$country) |> 
  select(country, pulmonary = new_sp_f2534, extrapulm = new_ep_f2534)

tb_amer
```

*Exercise 4.*

1.  Try making a bar chart to compare the total new cases of tuberculosis across countries. Each bar should be split in two different colours to indicate whether cases are pulmonary or extra-pulmonary TB.
2.  What issue are you facing, and how can you fix it?
3.  (Extra practice) Add population data from the `s_america` dataframe to `tb_amer`.
4.  (Test your memory) Add a new column with each country's incidence risk of tuberculosis (regardless of site) in 2010. (Incidence risk = number of new cases divided by number of people at risk over a given time frame.)
5.  (Extra practice) Create another bar chart, this time showing incidence, rather than new cases.
6.  (Challenge question) Bar charts are usually clearer if the bars are ordered by count, rather than alphabetically. Search online how to do this.

```{r}
ggplot(tb_amer, aes( # complete here
                    )) + 
  geom_bar(stat = "identity")
```

------------------------------------------------------------------------

## Pivot wider

Sometimes it might be helpful to go the other way - turn a long dataset wider, i.e., turning rows into columns. For example, to facilitate calculating new columns, or to increase readibility.

Let's look at the `meds` dataframe, which contains information on days of each antifungal in long format.

```{r}
meds
```

The function `pivot_wider()` from `dplyr` can do this.

We must specify at least two arguments:

1.  Which column contains the labels that will be turned into column names;
2.  Which column contains the values that will be sorted in each new column.

```{r}
meds |>
  pivot_wider(names_from = "antifungal", values_from = "days")
```

Now let's look at another subset of the WHO tuberculosis data. This time we're comparing new pulmonary TB cases among 35-44yo men in three European countries from 1995 to 2010.

```{r}
tb_eur <-  who |> 
  filter(!is.na(new_sp_m3544), iso3 %in% c("DEU", "FRA", "GBR", "ITA", "ESP"), year %in% seq(1995, 2010, 5)) |> 
  select(country = iso3, year, new_cases = new_sp_m3544)
 
tb_eur
```

*Exercise 5.*

1.  Reshape this dataset so that there is one column per year, containing the new number of cases in that country in that year.
2.  Change the name of the columns so that the numbers they contain are clear.

------------------------------------------------------------------------

## More resources

-   Joining data

    -   Epi R Handbook, Chapter 14: [Joining data](https://epirhandbook.com/en/joining-data.html)
    -   `dplyr` documentation, [Two-table verbs](https://dplyr.tidyverse.org/articles/two-table.html)
    -   R for Data Science, Chapter 13: [Relational data](https://r4ds.had.co.nz/relational-data.html)

-   Reshaping data

    -   Epi R Handbook, Chapter 12: [Pivoting data](https://epirhandbook.com/en/pivoting-data.html)
    -   `tidyr` documentation, [Pivoting vignette](https://tidyr.tidyverse.org/articles/pivot.html)
    -   R for Data Science, Chapter 12: [Tidy data](https://r4ds.had.co.nz/tidy-data.html)

------------------------------------------------------------------------

## Solutions

```{r ex1_solution}
# Trial
bind_rows(lab, site8)

# Fix
site8 |> rename(species = isolate) |> 
  bind_rows(lab)
```

```{r ex2_solution}
# Q1
lab_demographics<- left_join(lab, demogr, by = "id")

# Q2
lab_demographics|>
  ggplot(aes(age, bdg_0, shape = species)) + 
  geom_point()
```

```{r ex3_solution}
# Q1
# `lab` contained 10 rows, but the new dataset has to 15 rows. Some IDs are now duplicated, and the same happened with values in the BDG columns.

# Q2
# This is because `meds` contained more than 1 row per participant, if that participant received more than 1 antifungal drug. 
```

```{r ex4_solution}
# Q1
ggplot(tb_amer, aes(y = country, x = pulmonary)) + 
  geom_bar(stat = "identity")
# You can either plot pulmonary OR extrapulmonary cases, not both, because they're not in the same column

# Q2
tb_amer |>
  pivot_longer(-country, names_to = "site", values_to = "new_cases") |>
  ggplot(aes(y = country, x = new_cases, fill = site)) + 
  geom_bar(stat = "identity")

tb_amer |>
  # Q3
  left_join(s_america, by = "country") |>
  # Q4
  mutate(incidence_per100k = 100000 * (pulmonary + extrapulm) / popul) |> 
  ggplot(aes(
    # Q5
    x = incidence_per100k,
    # Q6
    y = forcats::fct_reorder(country, incidence_per100k)
    )) + 
  geom_bar(stat = "identity")
```

```{r ex5_solution}
# Q1
tb_eur |> pivot_wider(names_from = year,
                      values_from = new_cases,
                      # Q2
                      names_prefix = "new_TB_35-44M_")
```
