---
title: "8: Analysing continuous outcomes"
subtitle: "Introduction to R for health data"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
---

-------------------------------------------------------------------------------

# What you will learn

* Introduction to data analysis of continuous outcomes
 - data visualisation
 - t-test
 - simple linear regression

------------------------------------------------------------------------

# Recap from topic 7

Run this chunk to recreate some data from a UKHSA table on completed UK primary immunisations at 12 months by country (Apr - June 2022).

```{r ex_recap}
(vax <- tibble(country = c("Wales", "Northern Ireland", "Scotland", "England"),
       DTaP_IPV_Hib_HepB3 = c(94.9, 93.2, 96.1, 91.9),
       PCV1 = c(96.7, 94.9, 96.4, 94.1),
       Rota2 = c(93.6, 91.5, 94.5, 90.5),
       MenB2 = c(94.6, 93.4, 95.6, 92.0)
       ))

(auth <- tibble(area = c("England", "Scotland", "Wales", "Norther Ireland"),
               local_authorities = c(152, 14, 7, 4)))
```
Join the two tables into one.





-------------------------------------------------------------------------------

```{r}
library(magrittr)
library(tidyverse)
library(gtsummary)
```

-------------------------------------------------------------------------------


## Introduction

Let's load the Ambition dataset. This contains some simulated result of the AMBITION study, a clinical trial investigating the use of single-dose liposomal amphotericin B to treat cryptococcal meningitis, instead of the standard (longer) course.

Amphotericin B can damage the kidneys, so we're interested in the kidney function (creatinine) in these two groups.

```{r import_data}
amb <- rio::import("data/ambition.rds")
```

Just to familiarise yourself with the dataset:

* Exposure variable: `dose_ampho` (standard dosing versus single [low] dose)

* Outcome variable: `crea_2w` (serum creatinine level at 2 weeks)

* Baseline characteristics:
 - `age` (integer)
 - `sex` (binary)
 - `country` (categorical)
 - `crea_0` (baseline creatinine, in mmol/L - a measure of kidney function)
 - `egfr_0` (baseline eGFR - another measure of kidney function)

-------------------------------------------------------------------------------

## Average outcome value by a binary exposure

Let's imagine our research question is: "Do different doses of amphotericin lead to different serum creatinine level at 2 weeks?"

To have a first idea we can summarise the creatinine outcome in the two treatment arms, by calculating an average.

```{r}
amb |>
  group_by(dose_ampho) |>
  summarise(mean(crea_2w),
            sd(crea_2w))
```

The average creatinine at 2 weeks is 110 mmol/L in the group receiving standard dose, and 98 mmol/L among those who had low dose. In this sample, there's a difference of 12 mmol/L. How do we know if this reflects a true difference in the whole population (everyone who receives these drugs in these circumstances), and is not only visible in our sample due to chance?

-------------------------------------------------------------------------------

## Visualising a continuous outcome by a binary exposure

Let's get more insight by visualising the data.

We have a categorical exposure variable (treatment arm) and a continuous outcome variable (creatinine at 2 weeks). Which of these graphs are more appropriate?

- bar chart
- Box plot
- histogram
- line chart
- scatter plot
- violin plot

*Exercise 1*

Use an appropriate graph to visualise the distribution of creatinine at 2 weeks according to treatment arm.

```{r ex1}

```

From your data visualisation, you can see that there is quite a lot of overlap between creatinine values in both groups, however a large portion of the interquartile ranges don't overlap.

This is helpful to understand the data better, but still doesn't tell us if this difference could be due to chance.

-------------------------------------------------------------------------------

## *t*-test

In order to do hypothesis testing on the difference between two means, we can use a *t*-test (subject to some assumptions).

The function for the t test is `t.test()`.

In R statistical tests, variables listed before the tilde `~` are the outcome/response/dependent variables, and variables after the tilde are the exposure/independent variables.

Our hypothesis is that creatinine at 2 weeks is different according the dose in LAmB.

Our *null* hypothesis is that there is no difference (difference in means = 0).

```{r}
# With magrittr pipe
amb %$% t.test(crea_2w ~ dose_ampho)

# Base R equivalent
# t.test(amb$crea_2w ~ amb$dose_ampho)
```

The key outputs here are:  
1. the p-value, <0.001. This means that there is very strong evidence against the null hypothesis, and so very strong evidence that creatinine at 2 weeks is indeed different according to treatment group.
2. the means in the two groups - from which we get the direction and the magnitude of the effect. In this case, this means that on average, patients who received single-dose LAmB have a creatinine that's 8.3-15.0 units lower than those in the control group.

NB: By default, R uses a Welch t-test, and not a Student t-test. Welch is generally better because it doesn't assume that the variances in the two groups are equal.

*Exercise 2.*

Now try assessing with a t-test whether *baseline* creatinine is linked to female or male sex.

```{r ex2}

```

How do you interpret these results?


*Exercise 3.* 

Assess whether baseline eGFR is different in people from Uganda compared to people from Malawi.

Hint: the "country" variable contains more than 2 countries. How do restrict the dataset only to people from Uganda and Malawi?
```{r}

```

(What about if you have more than 2 groups in your exposure variable? You can't use a t test, you need to use ANOVA.)

-------------------------------------------------------------------------------

## A continuous outcome by a continuous exposure

Now let's go through another research question: "Is creatinine at baseline related to age, among patients with cryptococcal meningitis?"

Both variables are continuous now, so we can't calculate mean creatinine according to age unless we categorised the age group. What we can do is plotting this.

*Exercise 4.* 

Plot creatinine at baseline against age. Do you think there's a relationship?

```{r}

```

-------------------------------------------------------------------------------

## Linear regression

Linear regression is used to conduct hypothesis testing with a continuous outcome and one or more exposure variables. If there's just one binary exposure variable, it will be equivalent to a t-test.

What's great about linear regression is that we can visually represent it. In `ggplot2`, we can do this with `geom_smooth()`:

```{r}
amb |> ggplot(aes(x = age, y = crea_0)) +
  geom_point() +
  geom_smooth(method = "lm") # lm stands for linear (regression) model
```

What do you think of this regression line? Is there a relationship between creatinine and age?

The null hypothesis in linear regression is that in the population, this regression line is flat: in other words, there is no relationship between our exposure (age) and our outcome (creatinine). This is numerically equivalent to saying that the slope of that line is 0.

The alternative hypothesis is that the slope is somewhat different to 0.

The function is `lm()` (linear model) and its syntax is, annoyingly, different than `t.test()`. The dataset is specified as the "data" argument. We can then pipe into the `summary()` function to get the output we need.
```{r}
linear_crea0_age <- lm(crea_0 ~ age,
                       data = amb)

summary(linear_crea0_age)

confint(linear_crea0_age)
```

We're interested in these key outputs:
- The intercept estimate (In this case, 55 mmol/L)
- the age coefficient estimate (In this case, 0.53)
- The p-value for the age estimate - visualised as Pr(>|t|) - in this case, <0.001.

From the intercept and the coefficient, we can write a regression equation:

creatinine = 0.53 * age + 55

This means that for every year of life, creatinine increases by 0.53 mmol/L.

How confident are we about this statement? Very confident: the p value is <0.001. We can say that there is very strong evidence against the null hypothesis.

(The intercept, 55, mathematically refers to the mean creatinine at age 0 years. But in this dataset the youngest person is 18, so that value is purely extrapolated by continuing the regression line to the left. It does not have any clinical meaning.)

*Exercise 5.* 

Now assess the relation between *eGFR at baseline* and *age*:  
1. Draw a scatterplot with a linear regression line. What do you think of the distribution?
2. Conduct hypothesis testing with a linear regression model.
3. Write the regression equation. What does this mean?
4. How confident are you that this results reflects the true population values?
```{r}
# 1: scatterplot


# 2: linear model


# 3: regression equation


# 4: interpretation


```

### Nice regression table

Remember package `gtsummary`? We used it in topic 3 to create automatic descriptive tables, with `tbl_summary()`. It also has `tbl_regression()` which helps make regression output nicer to read.

```{r}
tbl_regression(linear_crea0_age)
```


-------------------------------------------------------------------------------

*Challenge exercise 6.*

Creatinine at 2 weeks is not very meaningful - in real life you'd compare that value with the baseline creatinine, and this is what the trial is looking at.
1. Create a new variable that represents the percent change in creatinine.
2. Visualise the distribution of this change in the two treatment arms.
3. Use an appropriate statistical test to assess whether there's a relation between change in creatinine and treatment arm.
```{r}
# 1


# 2


# 3


```




---

Solutions

```{r ex_recap_solution}
# Option A
vax |>
  rename(area = country) |> 
  left_join(auth, by = "area")

# Option B (short-cut)
left_join(vax, auth, by = c("country" = "area"))
```

```{r ex1_solution}
# Option A: Box plot
ggplot(amb, aes(y = crea_2w, x = dose_ampho)) +
  geom_boxplot() +
  coord_fixed(ratio = 0.02)

# Option B: violin plot
ggplot(amb, aes(y = crea_2w, x = dose_ampho)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  coord_fixed(ratio = 0.02)

# Option C: stratified histogram
ggplot(amb, aes(x = crea_2w, fill = dose_ampho)) +
  geom_histogram(binwidth = 12.5) +
  facet_grid(rows = vars(dose_ampho))
```


```{r ex2_solution}
amb %$% t.test(crea_0 ~ sex)
```

```{r ex3_solution}
amb |>
  filter(country == "uganda" | country == "malawi") %$%
  t.test(egfr_0 ~ country)
```

```{r ex4_solution}
amb |> ggplot(aes(x = age, y = crea_0)) + geom_point()
```

```{r ex5_solution}
# 1: scatterplot
amb |> ggplot(aes(x = age, y = egfr_0)) +
  geom_point(colour = "darkgreen") +
  geom_smooth(method = "lm", colour = "purple")

# 2: linear model
lm(egfr_0 ~ age,
   data = amb) |> 
  summary()

# 3: regression equation
# eGFR = -1.34 * age + 129

# 4: interpretation
# p-value < 0.001. There's very strong evidence that eGFR decreases with age.
```

```{r ex6_solution}
# 1
amb <- amb |>  mutate(change_creat = 100*(crea_2w - crea_0)/crea_0)
amb |> select(ID, crea_0, crea_2w, change_creat)

# 2
amb |> ggplot(aes(dose_ampho, change_creat)) + geom_boxplot()

# 3, option A
amb %$% t.test(change_creat ~ dose_ampho)

# 3, option B
lin_creat_dose <- lm(change_creat ~ dose_ampho,
   data = amb)

gtsummary::tbl_regression(lin_creat_dose)
```

